---
  output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-"
)
```

# naflex: Flexible options for handling missing values

<!-- badges: start -->
[![R-CMD-check](https://github.com/dannyparsons/naflex/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/dannyparsons/naflex/actions/workflows/R-CMD-check.yaml)
[![codecov](https://codecov.io/gh/dannyparsons/naflex/branch/master/graph/badge.svg?token=MSQKXE5UYR)](https://codecov.io/gh/dannyparsons/naflex)
[![CRAN_Status_Badge](https://www.r-pkg.org/badges/version/naflex)](https://cran.r-project.org/package=naflex)
[![license](https://img.shields.io/badge/license-LGPL (>= 3)-lightgrey.svg)](https://www.gnu.org/licenses/lgpl-3.0.en.html)
<!-- badges: end -->

The `naflex` R package provides additional flexibility for handling missing values in summary functions beyond the existing options (`na.rm = TRUE`/`FALSE`) available in base R.

Most summary functions in base R e.g. `mean`, provide the two extreme options for handling missing values:

1. calculate the summary ignoring all missing values (`na.rm = TRUE`), or
2. require no missing values for the summary to be calculated (`na.rm = FALSE`)

In many cases, something in between these two extremes is often more appropriate. For example, you may wish to give a summary statistic if less than `5%` of values are missing.

`naflex` provides helper functions to facilitate this flexibility. It allows for omitting missing values conditionally, using four types of checks:

1. a maximum proportion of missing values allowed
2. a maximum number of missing values allowed
3. a maximum number of consecutive missing values allowed, and
4. a minimum number of non-missing values required.

## Installation

```{r eval = FALSE}
# Install the development version from GitHub:
devtools::install_github("dannyparsons/naflex")
```

## Usage

The main function in `naflex` is `na_omit_if`. 

When wrapped around a vector in a summary function, `na_omit_if` ensures that the summary value is calculated when the checks pass, and returns `NA` if not. The example below shows how to calculate the `mean` conditionally on the proportion of missing values.

```{r}
library(naflex)

x <- c(1, 3, NA, NA, 3, 2, NA, 5, 8, 7)

# Calculate if 30% or less missing values
mean(na_omit_if(x, prop = 0.3))
# Calculate if 20% or less missing values
mean(na_omit_if(x, prop = 0.2))
```

Four types of checks are available:

1. `prop`: the maximum proportion (0 to 1) of missing values allowed
2. `n`: the maximum number of missing values allowed
3. `consec`: the maximum number of consecutive missing values allowed, and
4. `n_non`: the minimum number of non-missing values **required**.

If multiple checks are specified, all checks must pass for missing values to be removed. For example, although there are less than 4 missing values in `x`, there are two consecutive missing values, hence the `consec = 1` check fails in the example below the result is `NA`.

```{r}
mean(na_omit_if(x, n = 4, consec = 1))
```
The use of `%>%` ("pipe") from `magrittr` can be used to make the code look clearer and more familiar. The beginning of the line is now the same as standard R and it moves `na_omit_if` after `x` which then appears more like an option within the function, like `na.rm`, which is how you might think about `na_omit_if` conceptually in this case.

```{r, warning = FALSE}
require(magrittr)

sum(x %>% na_omit_if(prop = 0.25))
```

Note that you should not use `na_omit_if` with `na.rm = TRUE` in the summary function since this will always remove missing values so the checks are essentially ignored.

## How `naflex` works & more details

`na_omit_if` works by removing the missing values from `x` if the checks pass, and leaves `x` unmodified otherwise.

```{r}
# Missing values removed
na_omit_if(x, n = 4)
```
`na_omit_if` can be thought of like an extension of `stats::na.omit` and if missing values are removed, an `na.action` attribute and `omit` class are added for consistency with `stats::na.omit`.

```{r}
# Missing values not removed, x is unmodified
na_omit_if(x, n = 2)
```
A further set of four `na_omit_if_*` functions are provided for doing the same thing but restricted to a single check e.g. `na_omit_if_n(x, 2)`.

`na_check` has the same parameters as `na_omit_if` but returns a logical indicating whether the checks pass. It is used internally in `na_omit_if` and may also be a useful helper function.

```{r}
if (na_check(x, n = 4, consec = 1)) print("NA checks pass") else ("NA checks fail")
```

A set a four `na_check_*` functions are also provided for doing the same thing restricted to a single check e.g. `na_check_prop(x, 0.2)`

Finally, `naflex` provides a set of helper functions for calculating missing value properties used in these checks.

```{r}
na_prop(x)
na_n(x)
na_consec(x)
na_non_na(x)
```

## Compared to base R

In base R, this functionality can often be achieved using a combination of `ifelse`, `is.na` and the option `na.rm = TRUE`.`naflex` aims to simplify, shorten and standardise this process for users.

For example, the equivalent of:

```{r}
mean(na_omit_if(x, n = 4, prop = 0.2))
```

in base R is:

```{r}
ifelse(sum(is.na(x)) <= 4 && mean(is.na(x)) <= 0.2, mean(x, na.rm = TRUE), NA)
```


## Relevance to applications

The World Meteorological Organization (WMO) Guidelines on the Calculation of Climate Normals<sup id="a1">[1](#f1)</sup> recommends that a monthly mean value calculated from daily values should be calculated as long as there are no more than `10` missing values and no more than `4` consecutive missing values in the month.

# References
<sup id="f1">1</sup> <a href="https://library.wmo.int/index.php?lvl=notice_display&id=20130#.XljKS84zZnI">WMO Guidelines on the Calculation of Climate Normals</a> [↩](#a1)
