---
  output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-"
)
```

# naflex: Flexible options for handling missing values

<!-- badges: start -->
[![R-CMD-check](https://github.com/dannyparsons/naflex/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/dannyparsons/naflex/actions/workflows/R-CMD-check.yaml)
[![codecov](https://codecov.io/gh/dannyparsons/naflex/branch/master/graph/badge.svg?token=MSQKXE5UYR)](https://codecov.io/gh/dannyparsons/naflex)
[![CRAN_Status_Badge](https://www.r-pkg.org/badges/version/naflex)](https://cran.r-project.org/package=naflex)
[![license](https://img.shields.io/badge/license-LGPL (>= 3)-lightgrey.svg)](https://choosealicense.com/)
<!-- badges: end -->

The `naflex` R package provides additional flexibility for handling missing values in summary functions beyond the existing (\code{na.rm = TRUE/FALSE}) options available in base R.

Most summary functions in base R e.g. `mean`, provide the two extreme options for handling missing values:

1. calculate the summary ignoring all missing values (`na.rm = TRUE`), or
2. require no missing values for the summary to be calculated (na.rm = FALSE`)

In many cases, something in between these two extremes is often more appropriate. For example, you may wish to give a summary statistic if less than `5%` of values are missing.

`naflex` provides helper functions to facilitate this flexibility. It allows for omitting missing values conditionally, using four types of requirements:
- a maximum proportion of missing values allowed
- a maximum number of missing values allowed
- a maximum number of consecutive missing values allowed, and
- a minimum number of non-missing values required.

## Installation

```{r eval = FALSE}
# Install the development version from GitHub:
# install.packages("devtools")
devtools::install_github("dannyparsons/naflex")
```

## Usage

The main function in `naflex` is `na_omit_if`. 

When wrapped around a vector in a summary function, `na_omit_if` ensures that the summary value is calculated when the checks pass, and returns `NA` if not. The example below shows  how to calculate the `mean` conditionally on the proportion of missing values.

```{r}
library(naflex)

x <- c(1, 3, NA, NA, 3, 2, NA, 5, 8, 7)

# Calculate if 4 or less missing values
mean(na_omit_if(x, prop = 0.3))
# Calculate if 2 or less missing values
mean(na_omit_if(x, prop = 0.2))
```

Four types of checks are available:

- `prop`: the maximum proportion (0 to 1) of missing values allowed
- `n`: the maximum number of missing values allowed
- `consec`: the maximum number of consecutive missing values allowed, and
- `n_non`: the minimum number of non-missing values **required**.

If multiple checks are specified, all checks must pass for missing values to be removed. For example, the code below returns `NA`. Although there are less than 4 missing values in `x`, there are two consecutive missing values, hence the `consec = 1` check fails.

```{r}
mean(na_omit_if(x, n = 4, consec = 1))
```
Note that you cannot use this option with `na.rm = TRUE` in the summary function since this will always remove missing values so the checks are essentially ignored.

## Details & How `naflex` works

`na_omit_if` works by removing the missing values from `x` if the checks pass, and leaving `x` unmodified otherwise.

```{r}
# Missing values removed
na_omit_if(x, n = 4)
```
If missing values are removed, an `na.action` attribute and `omit` class is added for consistency with `stats::na.omit`.

```{r}
# Missing values not removed
na_omit_if(x, n = 2)
```
A set of four `na_omit_if_*` functions are provided for doing the same thing restricted to a single check e.g. `na_omit_if_n(x, 2)`

`na_check` has the same parameters as `na_omit_if` but returns a logical indicating whether the checks pass. It is used internally in `na_omit_if` but may also be a useful helper function. 

```{r}
if (na_check(x, n = 4, consec = 1)) print("NA checks pass") else ("NA checks fail")
```

A set a four `na_check_*` functions are also provided for doing the same thing restricted to a single check e.g. `na_check_prop(x, 0.2)`

Finally, `naflex` provides a set of helper functions for calculating missing value properties used in these checks.

```{r}
na_prop(x)
na_n(x)
na_consec(x)
na_non_na(x)
```

## Compared to base R

You can do this by first calculating the proportion of missing values and then using `na.rm = TRUE` as appropriate. `naflex` aims to simplify this process.

## Relevance to applications

The World Meteorological Organization (WMO) Guidelines on the Calculation of Climate Normals<sup id="a1">[1](#f1)</sup> recommends that a monthly mean value calculated from daily values should be calculated as long as there are no more than `10` missing values and no more than `4` consecutive missing values in the month.

# References
<sup id="f1">1</sup> <a href="https://library.wmo.int/index.php?lvl=notice_display&id=20130#.XljKS84zZnI">WMO Guidelines on the Calculation of Climate Normals</a> [↩](#a1)
