---
  output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-"
)
```

# naflex: Flexible options for handling missing values

<!-- badges: start -->
[![R-CMD-check](https://github.com/dannyparsons/naflex/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/dannyparsons/naflex/actions/workflows/R-CMD-check.yaml)
[![codecov](https://codecov.io/gh/dannyparsons/naflex/branch/master/graph/badge.svg?token=MSQKXE5UYR)](https://codecov.io/gh/dannyparsons/naflex)
[![CRAN_Status_Badge](https://www.r-pkg.org/badges/version/naflex)](https://cran.r-project.org/package=naflex)
[![license](https://img.shields.io/badge/license-LGPL (>= 3)-lightgrey.svg)](https://choosealicense.com/)
<!-- badges: end -->

The `naflex` R package provides additional flexibility for handling missing values in summary functions beyond the existing (\code{na.rm = TRUE/FALSE}) options available in base R.

Most summary functions in base R e.g. `mean`, provide the two extreme options for handling missing values:

1. calculate the summary ignoring all missing values (`na.rm = TRUE`), or
2. require no missing values for the summary to be calculated (na.rm = FALSE`)

In many cases, something in between these two extremes is often more appropriate. For example, you may wish to give a summary statistic if less than `5%` of values are missing.

`naflex` provides helper functions to facilitate this flexibility. It allows for omitting missing values conditionally, using four types of requirements:
- a maximum proportion of missing values allowed
- a maximum number of missing values allowed
- a maximum number of consecutive missing values allowed, and
- a minimum number of non-missing values required.

## Installation

```{r eval = FALSE}
# Install the development version from GitHub:
# install.packages("devtools")
devtools::install_github("dannyparsons/naflex")
```

## Usage

The main function in `naflex` is `na_omit_if`. When wrapped around a vector inside a summary function, it ensures the summary function returns 

## Compared to base R

You can do this by first calculating the proportion of missing values and then using `na.rm = TRUE` as appropriate. `naflex` aims to simplify this process.

## Relevance to applications

The World Meteorological Organization (WMO) Guidelines on the Calculation of Climate Normals<sup id="a1">[1](#f1)</sup> recommends that a monthly mean value calculated from daily values should be calculated as long as there are no more than `10` missing values and no more than `4` consecutive missing values in the month.

# References
<sup id="f1">1</sup> <a href="https://library.wmo.int/index.php?lvl=notice_display&id=20130#.XljKS84zZnI">WMO Guidelines on the Calculation of Climate Normals</a> [↩](#a1)
